---
title: "Land Use and Water Quality"
author:
- "Dr. Matthew Ross"
- 'matt.ross(at)colostate.edu'
- 'Office Hours: M,TR 9:00-10:00 AM'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
---


# (PART) Syllabus {-}

# Land Use and Water Quality (WR418)


This working repository hosts all lecture material, code, assignments, and the syllabus for Land Use and Water Quality (WR418) taught at Colorado State University.

The long-term vision for this project is to work towards an open access book
similar to [R for Data Science](https://r4ds.had.co.nz/), which can be used to teach
college students about the interaction of human land-use and downstream water quality
change.

## Course (and eventual book) Goals

The broadest goals of the book/course are for students to develop both an 
intuitive and quantitative understanding of:

- How human decisions about land-use alters how water moves through the landscape,
 the materials it picks up, and, what we can and can't do to improve degraded water quality.
 
- The chemical reactions and processes that govern the connection between land-use
and downstream water quality, over short and long time-scales.
 
- The history of human impacts to water quality from water quality impacts in 
ancient rome to modern pharmaceuticals entering our lakes and rivers.

- The status of water quality in lakes and rivers across the globe and how it relates
to legislation, developoment, and environmental change broadly.

# Approach

To reach these educational goals we will use a variety of approaches focused on 
engaging students directly with primarily literature on water quality, discussing
current events related to land-use and water quality, quantitative analyses of 
land-use and water quality change over time, and a focused final project. 

## Assignments

All assignments will be distributed through canvas

## Notes
If you learn by taking notes, I encourage you to take notes in class when there 
are lectures. However, all lectures will be posted online on my [GitHub account](https://github.com/OpenEcoDataLab/Land_Use_Water_Quality-WR418). Additionally,
a more verbose version of the notes will be on [this site](http://matthewrvross.com/wr418/land-use-and-water-quality-wr418.html) (look to your left).

## Attendance
I will not take attendance, but this class will heavily depend on you being present. There
will be a participation grade that will reflect a combination of your consistent presence, focus in class, and participation in group discussions (it's okay if you don't like speaking in front of the whole class, but you should particpate during breakout period). 

## Grading 

```{r, echo=F}
library(knitr)
library(kableExtra)
library(tibble)
library(lubridate)

grading <- tibble(Assignments='50%',
                  Midterm='15%',
                  Final='20%',
                  Participation='15%')

kable(grading)
```

## Schedule 
This schedule will likely change considerably as we move through the course,
but I will keep it updated. Every Friday there will be an assignment due the next 
Friday, unless otherwise stated.

```{r, echo=F}

schedule <- tibble(`Week Number`=1:16,
                   `Week Starting`=seq(ymd('2018-01-21'),ymd('2018-05-10'),by=7),
                   Topic=c('Getting started and course review',
                           'Global history of water quality change, point source pollution',
                           'Non-point source pollution, nutrients, sediment, dissolved organic carbon',
                           'Algae, eutrophication, cyanobacterial blooms',
                           'weathering reactions, cations, anions',
                           'Water treatment approaches and history,MIDTERM',
                           'Metals chemistry, pH as a master variable',
                           'Organic contaminants',
                           'Legislation governing water quality in the USA and abroad',
                           'Tracing water quality problems with isotopes and sensors',
                           'Remote sensing of land-use change and water quality',
                           'River/Lake network water quality influences',
                           'Mapping water quality issues in Colorado and Fort Collins',
                           'Water quality legacy impacts, biotic indices',
                           'Contaminant interactions, phytolosis, transformation',
                           'BYO contaminant presentations (final)'))
kable(schedule) %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Academic integrity
Academic integrity: You are responsible for adhering to all university policies on academic
integrity and student conduct https://tilt.colostate.edu/integrity/knowTheCode/. TILT
has a number of resources for students related to writing and study skills:
https://tilt.colostate.edu/integrity/resourcesStudents/.