---
title: "Land Use and Water Quality"
author:
- "Dr. Matthew Ross"
- 'matt.ross(at)colostate.edu'
- 'Office Hours: M,TR 10:00-11:00 AM'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
---


# (PART) Syllabus {-}

# Land Use and Water Quality (WR418)


This working webpage hosts all lecture material, code, assignments, and the syllabus for Land Use and Water Quality (WR418) taught at Colorado State University. This class is undergoing
rapid change and 

The long-term vision for this project is to work towards an open access book
similar to [R for Data Science](https://r4ds.had.co.nz/), which can be used to teach
college students about the interaction of human land-use and downstream water quality
change.

## Course Goals

The broadest goals of the course are for students to develop both an 
intuitive and quantitative understanding of:

- How human decisions about land-use alters how water moves through the landscape,
 the materials it picks up, and, what we can and can't do to improve degraded water quality.
 
- The chemical reactions and processes that govern the connection between land-use
and downstream water quality, over short and long time-scales.

- The status of water quality in lakes and rivers across the globe and how it relates
to legislation, developoment, and environmental change broadly.

# Approach

To reach these goals, we will have a dynamic classroom that incorporates lectures,
readings, discussion, presentation, work days, and more. Detailed approach notes 
and details are below. 

## Class policies

### Inclusive classroom

We will have an inclusive classroom: 
respecting and valuing the diverse backgrounds, perspectives, and identities
of individuals in the class. Students are expected to have an awareness of and
sensitivity to language or actions that may be exclusionary or alienating. I 
am committed to fostering a collaborative and inclusive class and appreciate
any feedback on how to improve my own practice.  

This is particularly important in a class that will have group work,
peer grading, and generally serve individuals with a large range in background
with watershed, chemistry, statistics, and other skills we will use in class. 
We all get better if we work together. 
A final note on treatment of others: one of the most important things when 
interacting with others who have more or less research experience than you is to
learn how to ask effective questions that show you understand and value 
other people's time. "What work
can I do to prepare my question in a way that makes it easiest for someone to 
help me?" I want you to ask questions and work collaboratively and do so in a way
that helps you and everyone else.


### Assignments

All assignments will be distributed through Canvas.

### Asking for help and code on the internet:

One of the most common ways people learn to code is to use the age-old 
(okay, maybe 10 years) technique of asking the internet for help. This is a great idea!
But! When you do you should ask for help in specific ways that enable people
to answer your question more easily and clearly. Great instructions are here: 
https://blog.revolutionanalytics.com/2014/01/how-to-ask-for-r-help.html
and a more general guide to help in R is here:
https://www.r-project.org/help.html.

### Using other people's code

Inevitably, you will want to perform a task in R that someone else has already 
done. This is useful and part of why the R community is so great. But! If 
you use other people's code, you should:

  a) explitly cite where you got the code or the inspiration for it
  
  b) work hard to understand the code and what it does, break it down into 
  pieces, and try to rebuild it 
  
Code that you did not write and is not cited will be treated as academic 
plagiarism. Assignments will vary in how much code is allowed to be shared 
between students, but the general rule is that you should be submitting your own
code or code from your own team. As an example of best practices, a lot of the
inspiration for this page (and the class in general) comes from Mine Cetikaya-Rundel
and her amazing [STA 199](http://www2.stat.duke.edu/courses/Spring18/Sta199/policies/) 
course at Duke

## Academic integrity

Academic integrity: You are responsible for adhering to all university policies on academic
integrity and student conduct https://tilt.colostate.edu/integrity/knowTheCode/. TILT
has a number of resources for students related to writing and study skills:
https://tilt.colostate.edu/integrity/resourcesStudents/.

## Attendance

This class will heavily depend on you being present.
There will be a participation grade that will reflect a combination of 
your consistent presence, focus in class, and participation in group work. We 
will also occasionally have quizzes to check in on your understanding of material
but also to make sure you are present. 

## Notes

If you learn by taking notes, I encourage you to take notes in class when there 
are lectures. However, all lectures will be posted online on my [GitHub account](https://github.com/OpenEcoDataLab/Land_Use_Water_Quality-WR418). Additionally,
a more verbose version of the notes will be on [this site](http://matthewrvross.com/wr418/land-use-and-water-quality-wr418.html) (look to your left).


## Grading 

```{r, echo=F,message=F,warning=F}
library(knitr)
library(kableExtra)
library(tibble)
library(lubridate)
library(googlesheets4)


grading <- tibble(Assignments='50%',
                  Midterm='20%',
                  Final='20%',
                  Participation='10%')

kable(grading)
```

## Schedule 

This schedule will likely change considerably as we move through the course,
but I will keep it updated. Every Friday there will be an assignment due the next 
Friday, unless otherwise stated.

```{r, echo=F}

schedule <- sheets_find('WR418, spring 2020') %>%
  dplyr::pull(id) %>%
  read_sheet() %>%
  dplyr::select(1:3)



kable(schedule) %>%
 kable_styling(bootstrap_options = c("striped", "hover", 
                                     "condensed", "responsive")) %>%
  column_spec(3,width='15em')
```

